# Data Transformation 4: Fine-tuned Data Manipulation 

## Intended Learning Outcomes {-}

The whole purpose of this lecture is to expose you to a couple of useful functions that would make your daily life easier when dealing with your own data. By the end of today, you will know

1. How to rename column headers
2. How to recode cell values based on conditions
3. How to obtain unique values from your data

This lesson is led by [Gaby Mahrholz](http://www.psy.gla.ac.uk/staff/index.php?id=GM005).

## Pre-steps

Before we can start to focus on some other functions that might come in handy for data manipulation, we need to make sure to load `tidyverse` into the library. 

```{r L6, message=FALSE}
library(tidyverse)
```


## rename()

`rename()` is a very useful function if we wanted to change column names. All available column are retained, so nothing is lost. If you had 5 variables in a tibble, and wanted to change the name of one of them, your output would be 5 columns in total - one column with a changed name and 4 columns with the previous names. The `rename()` function follows a very simple pattern of

```{r, eval=FALSE}
rename(data, new_column_name = old_column_name)
```

You can rename multiple column headers by including more arguments, separating them with a comma. Let's take a look at a simple example in `table1`.

```{r}
table1
```

If we wanted to change the column headers of `country` to `Country` and `population` to `Population`, and keep the columns `year` and `cases` as they are, we would code

```{r}
rename(table1, Country = country,
       Population = population)
```

If you want to select a few columns and rename them in the process, you could use `select()`. The structure is the same as with `rename()`, the only difference is that the new data only retains the columns mentioned explicitly within `select()`.

```{r, eval=FALSE}
select(data, new_column_name = old_column_name)
```

Modifying the `rename()` example from above would lead to selecting **only** the columns `country` and `population` from `table1` and renaming them as `Country` and `Population` at the same time.

```{r}
select(table1, Country = country,
       Population = population)
```


## Conditional statements

Conditional statements are used when you want to create an output value that is conditioned on an evaluation. Here, we are showing you how to use them in combination with `mutate()` by either modifying the values within one column, or by adding a new column.

Example scenario: Imagine that we have data in which a column `Sex` holds values of `1` and `2` that stand for `Male`, and `Female` participants respectively. Conditional statements make an evaluation: *if `Sex` is `1`, then put the value of `Male` in a new column called `Sex_char`; if `Sex` is `2` put `Female` in column `Sex_char`.*

Let's create some fake data for this. 40 participants (17 male, 20 female, 3 other), they are between 16 and 40 years of age, they study at 4 different institutions (University of Glasgow, University of Strathclyde, Glasgow Caledonian University, University of the West of Scotland), and we had them complete an R assignment and recorded their grade points.

```{r fake_data}
set.seed(999) #so we all get the same random data
fake_data <- tibble(PP_ID = 1:40,
                    Sex = c(rep(1,17), rep(2,20), rep(3,3)),
                    Age = sample(16:40, 40, replace = TRUE),
                    Institution = sample(c(rep(1,10), rep(2,8), rep(3,10), rep(4,10), rep(99,2))),
                    Grade_points = sample(c(0,9:22), 40, replace = TRUE))
glimpse(fake_data)

```

```{r}
kable(fake_data)
```


Now, the data doesn't look very readable at the moment. If we pass this data set onto a fellow researcher, they would ask us how the columns `Sex` and `Institution` are coded.

Best would be to adjust the values to make that clearer. There are the options of using the function `recode()`, an `ifelse` statement or `case_when()`.


### recode()

Can be used for character, numeric, and factors columns.

The structure for `recode()` is:

```{r, eval=FALSE}
recode(variable_to_recode, old_value = new_value)
```

This is a bit tricky to remember as it doesn't follow the usual tidyverse logic. In `select()`, `rename()`, and `mutate()` for example it is the other way around `new_thing = old_thing` or `new_thing = something_computational`)

In combination with `mutate()`, the structure looks as follows:

```{r, eval=FALSE}
mutate(data, new_column_name = recode(variable_to_recode, old_value = new_value))
```

What is a bit unusual in this structure is that we need a new column name within the `mutate()` function but also a reference column within `recode()` on which the conditional recoding is based upon.

The new column name within `mutate()` can either be an entirely new column name, or if it is an existing column name that original column would be overwritten by the new values we recode.

So, let's turn to our `fake_data`, and apply that logic for recoding the values of the column `Sex` where 1 = "Male", 2 = "Female", and 3 = "Other". We will add a new column `Sex_char` onto our `fake_data`.

The only tricky thing to remember here is that recoding numeric values requires backticks (if the numbers are on the left side of the equation).

```{r}
fake_data <- mutate(fake_data, Sex_char = recode(Sex, 
                                                 `1` = "Male", 
                                                 `2` = "Female", 
                                                 `3` = "Other"))
glimpse(fake_data)
```

```{block, type="warning"}
If you forget to recode one condition (for example ``\`3\` = "Other"``), values for 3 would be displayed as `NA` in the new column. This becomes more important when you decide to overwrite an existing column.
```



```{block, type="task"}
**Your turn**

The next variable to recode is `Institution`:

* 1 = UoG, 
* 2 = Strathclyde, 
* 3 = GCU, 
* 4 = UWS

In the questionnaire, two participants forgot to indicate which university they are studying at (coded as 99), but fortunately we know that they are at the University of Strathclyde (the test was distributed to 10 students at each institution, however, Strathclyde only has 8 responses).

Add a new column `Institution_char` to `fake_data` that holds the recoded values as
```


`r hide("Solution")`
```{r}
fake_data <- mutate(fake_data, Institution_char = recode(Institution, 
                                                         `1` = "UoG", 
                                                         `2` = "Strathclyde", 
                                                         `3` = "GCU", 
                                                         `4` = "UWS", 
                                                         `99` = "Strathclyde"))
glimpse(fake_data)
```
`r unhide()`

An alternative to `recode()` is `ifelse()`. Let's see how `Sex` and `Institution` could have been recoded using `ifelse()`.



### ifelse()

`ifelse()` is a good option when dealing with character or numeric input but struggles with dates and factors. None of the columns in `fake_data` are dates or factors, so we can easily use `ifelse()` for recoding variables.

The generic structure for `ifelse()` is:

```{r, eval=FALSE}
ifelse(test, yes, no)
```

* test: is an expression to test that creates an output of `TRUE` or `FALSE`
* yes: if test has a `TRUE` outcome, this is what would be coded in the new column
* no: if test has a `FALSE` outcome, this is what would be coded in the new column

The interpretation of `ifelse()` is: if the test expression is `TRUE`, write the value of yes, otherwise no.

In combination with `mutate()`, the structure looks as follows:

```{r, eval=FALSE}
mutate(data, new_column_name = ifelse(test, yes, no))
```

How would we apply that to recoding `Sex`? We would want the expression to test whether the Sex column holds a value of 1, then a new column should code it as "Male", otherwise as "Female". Let's add a new column `Sex_ifelse` to `fake_data`

```{r}
fake_data <- mutate(fake_data, Sex_ifelse = ifelse(Sex == 1, "Male", "Female"))
glimpse(fake_data)
```

AHHHH!!! But what about the 3 participants that chose option 3? They are now coded as "Female". Hmmm. One way to get around that would be a nested `ifelse()` statement. Instead of replacing the no argument (`FALSE` evaluation) with "Female", we could include a second `ifelse()` statement.

```{r}
fake_data <- mutate(fake_data, Sex_ifelse = ifelse(Sex == 1, "Male",
                                                   ifelse(Sex == 2, "Female", "Other")))
```

Now everything is coded as it should be.


```{block, type="task"}
**Your turn**

Recode `Institution` using `ifelse()`. Add a new column `Institution_ifelse` to `fake_data`. A quick reminder of the `Institution` values:

* 1 = "UoG", 
* 2 and 99 = "Strathclyde", 
* 3 = "GCU", and
* 4 = "UWS"
```


`r hide("Solution")`
```{r}
fake_data <- mutate(fake_data, Institution_ifelse = ifelse(Institution == 1, "UoG", 
                                                           ifelse(Institution == 3, "GCU", 
                                                                  ifelse(Institution ==  4, "UWS", "Strathclyde"))))

glimpse(fake_data)
```
`r unhide()`


It does look a bit convoluted, and there is definitely the need to count opening and closing brackets, but it gets us the output we wanted. Before showing you how this can be coded in a slightly neater way with `case_when()`, I'd like to highlight one advantage `ifelse()` has over `recode()`, namely that the test expression can evaluate more complex "things" as long as they return a logical value. For example, if we wanted to add a column `Pass_Fail` to `fake_data`, that holds values of "Pass" for grade points above 9, and "Fail" for values of 9 and below, we would code


```{r}
fake_data <- mutate(fake_data, Pass_Fail = ifelse(Grade_points > 9, "Pass", "Fail"))
glimpse(fake_data)
```

We could also add a column `Grade` to `fake_data` that tells us the primary band the students got in their test. `Grade_points` between 18 and 22 result in A, 15 to 17 = B, 12 to 14 = C, 9 to 11 = D, 6 to 8 = E, 3 to 5 = F, 1 and 2 = G, and 0 = H. Here we can use the `%in%` operator.

```{r}
fake_data <- mutate(fake_data, Grade = ifelse(Grade_points %in% 18:22, "A",
                                              ifelse(Grade_points %in% 15:17, "B",
                                                     ifelse(Grade_points %in% 12:14, "C",
                                                            ifelse(Grade_points %in% 9:11, "D",
                                                                   ifelse(Grade_points %in% 6:8, "E",
                                                                          ifelse(Grade_points %in% 3:5, "F",
                                                                                 ifelse(Grade_points %in% 1:2, "G","H"))))))))

glimpse(fake_data)
```


### case_when()

`case_when()` vectorises multiple `ifelse()` statements, and might potentially be a touch easier to digest visually than `ifelse()`.

There are few guidelines to follow when using `case_when()`:

* A sequence of two-sided formulas is needed. 
* The left hand side (LHS) determines which values match this case. The right hand side (RHS) provides the replacement value.
* The LHS must evaluate to a logical vector. The RHS does not need to be logical, but all RHSs must evaluate to the same type of vector.

Let's apply this to recoding the values in the column `Institution`. Combining `case_when()` with `mutate()`, we will add another column called `Institution_case_when` to `fake_data`.

```{r}
fake_data <- mutate(fake_data, Institution_case_when = case_when(
  Institution == 1 ~ "UoG",
  Institution == 3 ~ "GCU",
  Institution == 4 ~ "UWS",
  TRUE ~ "Strathclyde"
  )
)
glimpse(fake_data)
```

The `TRUE` statement at the end is similar to the the "else" in the `ifelse()` function. A further similarity between the two is that arguments are evaluated in order, so you must proceed from the most specific to the most general.


```{block, type="task"}
**Your turn**

Add a column `Grade_case_when` into `fake_data` using a combination of `mutate()` and `case_when()`, that recodes grade points to primary bands. Quick reminder:

* grade points 18 - 22 = A,
* 15 to 17 = B, 
* 12 to 14 = C, 
* 9 to 11 = D, 
* 6 to 8 = E, 
* 3 to 5 = F, 
* 1 and 2 = G, and 
* 0 = H.
```


`r hide("Solution")`
```{r}
fake_data <- mutate(fake_data, Grade_case_when = case_when(
  Grade_points %in% 18:22 ~ "A",
  Grade_points %in% 15:17 ~ "B",
  Grade_points %in% 12:14 ~ "C",
  Grade_points %in% 9:11 ~ "D",
  Grade_points %in% 6:8 ~ "E",
  Grade_points %in% 3:5 ~ "F",
  Grade_points %in% 1:2 ~ "G",
  TRUE ~ "H"
  )
)
glimpse(fake_data)
```
`r unhide()`


## distinct()

The `distinct()` function is used to remove duplicate rows in your dataframe. If we are using `distinct()` without specifying any arguments, R checks the whole row and excludes any row that is exactly repeated.

```{r}
fake_data_no_dup <- fake_data %>% 
  distinct()
```

We can see that all rows are still there. This is not surprising, given every participant has a unique ID number. However, `distinct()` does take arguments as well, namely the variable columns you are trying to determine uniqueness for. Say we we wanted to see how many different institutions we have data from.


```{r}
institution_distinct <- fake_data %>% 
  distinct(Institution_char)
institution_distinct
```

We can see that there are 4 distinct institutions. But what happened to the output? We still have the output from the column we specified but all other information is gone. One quick solution would be to use the argument `.keep_all` and set that to `TRUE` (the default here is `FALSE`).


```{r}
institution_distinct <- fake_data %>% 
  distinct(Institution_char, .keep_all = TRUE)
institution_distinct
```

In the new `institution_distinct` dataframe, we still have 4 institutions listed but we kept all 13 columns. If there are multiple rows with the same input (in this case 40 participants from 4 different institutions), `distinct()` only preserves the first row a new value occurs.

The `distinct()` function also allows you to eliminate duplicate rows with multiple variables in the dataframe. So if we wanted to see where participants were studying but also the different sexes taking part in the study, we would code...

```{r}
institution_sex_distinct <- fake_data %>% 
  distinct(Institution_char, Sex_char)
institution_sex_distinct
```

`institution_sex_distinct` has 10 observations, so that means, we have at least 1 male and 1 female participant from each of the 4 institutions. Additionally, Strathclyde and Glasgow have participants classifying as neither male nor female (here coded as "Other"). 

```{block, type="question"}
You can see they are not sorted by either of the column. Why is that?
```

`r hide("Solution")`
```{block, type="solved"}
The values are sorted by occurrence. R goes through the tibble row by row to determine whether the values are distinct from the ones in the row before. So if you look at your original dataframe `fake_data`, you can see the first row is a male participant from Glasgow Caledonian University. The participant in row 2 is from the University of Glasgow, but so is participant 3 (which means he is not listed in the distinct tibble). The next unique value would be detected in row 4, then in row 7, then in row 18, 20, 22, 23, 38, and 39.
```
`r unhide()`



## Recap

Let's just have some fun with `fake_data` whilst recapping a little bit of what we did in the past 5 weeks.

### Question 1

Select the variables `PP_ID`, `Sex_char`, `Age`, `Institution_char`, `Grade_points`, `Grade`, and `Pass_Fail` from `fake_data`and save it as `fake_data` in your Global Environment (i.e. overwrite/replace `fake_data`). Rename the following variables in the process:

* `PP_ID` as `Participant_ID`
* `Sex_char` as `Sex`
* `Institution_char` as `Institution`

```{r, eval=FALSE}
fake_data <- NULL
```


`r hide("Solution")`
```{r}
# solution using select:
fake_data <- fake_data %>% 
  select(Participant_ID = PP_ID, Sex = Sex_char, Age, Institution = Institution_char, Grade_points, Grade, Pass_Fail)

fake_data

# solution involving select and rename:
# fake_data <- fake_data %>% 
#   select(PP_ID, Sex_char, Age, Institution_char, Grade_points, Grade, Pass_Fail) %>%
#   rename(Participant_ID = PP_ID, Sex = Sex_char, Institution = Institution_char)

```

`r unhide()`



### Question 2

Calculate the average grade points for each University. Store the output as a new object `grade_points` in your `Global Environment`. The new column should be named `Average_grade_points`.

```{r, eval=FALSE}
grade_points <- NULL
```


`r hide("Solution")`
```{r}
grade_points <- fake_data %>% 
  group_by(Institution) %>% 
  summarise(Average_grade_points = mean(Grade_points)) %>% 
  ungroup()

grade_points
```

`r unhide()`


### Question 3

Join the `fake_data` and `grade_points`, and calculate how much each person differs from the mean of their university. Results should be stored in a column named `grade_diff`. Store the output in your `Global Environment` as a new object called `fake_data_meandiff`. Sort the data by `Institution` and `Sex`.

```{r, eval=FALSE}
fake_data_meandiff <- NULL
```


`r hide("Solution")`
```{r}
fake_data_meandiff <- left_join(fake_data, grade_points) %>% 
  mutate(grade_diff = Grade_points - Average_grade_points) %>% 
  arrange(Institution, Sex)

fake_data_meandiff
```

`r unhide()`



### Question 4

Our colleague is interested in what the grade point average for each university would be if we were only looking at the students who passed the R assignment initially (i.e. grade points above 9). Store the output in your `Global Environment` as a new object called `fake_data_pass`.

```{r, eval=FALSE}
fake_data_pass <- NULL
```

`r hide("Solution")`
```{r}
# filter Grade_points
fake_data_pass <- fake_data %>% 
  filter(Grade_points > 9) %>% 
  group_by(Institution) %>% 
  summarise(Average_grade_points_pass = mean(Grade_points)) %>% 
  ungroup()

# filter Pass_Fail
fake_data_pass <- fake_data %>% 
  filter(Pass_Fail == "Pass") %>% 
  group_by(Institution) %>% 
  summarise(Average_grade_points_pass = mean(Grade_points)) %>% 
  ungroup()

fake_data_pass
```

`r unhide()`



### Question 5

Now our colleague wants to know whether the males or the females did better at each university. They are asking for a table with 4 columns in which the first column titled `Institution` lists the university, the second column is called `Males` for the average grade point scores of the males at each institution, the third column called `Females` with average grade point scores for the females, and the fourth column `Better` should say `"female"` or `"male"` if the females or males had higher grade points respectively. Store the output in your `Global Environment` as a new object called `fake_data_sex`. Use `fake_data` as the reference dataframe.

```{r, eval=FALSE}
fake_data_sex <- NULL
```


`r hide("Solution")`
```{r}
fake_data_sex <- fake_data %>% 
  filter(Sex != "Other") %>% # this step could actually be left out
  group_by(Institution, Sex) %>% 
  summarise(Average_gp = mean(Grade_points)) %>% 
  ungroup() %>% 
  spread(Sex, Average_gp) %>% 
  select(Institution, Males = Male, Females = Female) %>% 
  mutate(Better = ifelse(Males > Females, "male", "female"))

fake_data_sex
```

`r unhide()`



## Summative Homework

The third summative assignment is available on moodle now.

Good luck.

Check that your Rmd file knits into a html file before submitting. Upload your Rmd file (not the knitted html) to moodle. 