# Data Visualisation 1


```{r warning=FALSE, echo=FALSE}
library(tidyverse)
```


##Introduction to the data

This data is provided freely from the Scottish Government, and it introduces the flow of people at different ages and sex into and out of the Scotland. We have several variables to work with:

FeatureCode =  codes given by the Scottish Governemnt
  DateCode = Year of data collected
  Measurement = what type of measurment it is (Here we have only counts)
  Units = units(here we have only people)
  Value = The actual counts
  Age = Age of the counted people (Separate age, and total)
  Sex = sex of the counted people (separate sex and total)
  `Migration Source` = Where or from the people are comming from ( Overseas, RUK)
  `Migration Type` = whether people are comming or leavin ( In or OUT and Net)


```{r}
migration_scot <- NULL 

```

The data is a bit messy so lets clean it a little bit before we can work with it. 

### Some warm-up tasks 

1. Keep only the total number of people by age ('All')
2. Keep only the necessary variable - DateCode, Value, Sex, Age, 'Migration Source' & 'Migration Type'
3. Filter the data , and keep only 2016
4. Keep only the information about the number of people that have came in and left scotland ( we do not want to know about the net value)




```{r}

traffic_scot <- NULL

```




## The package GGPLOT2


We use the ggplot2 package due to its ease of layering and variety. 

The ggplot2 package allows us to create beautiful and various plots, that can show us much more than just the simplicities of the base R plots. 





## Layering

Plots created with ggplot work by using layers. 

In the simplest form we have two layers. 

 Layer1: what data are we going to use
 Layer2: how do we want it to look

Layer 1: ggplot(data, aes(x,y, ...)) + 
Layer 2:   geom_something(...)



### Layer 1

Plots always start with *ggplot()*. This initialises the plot. It is used to tell *R* what data we are going to use. We can also set the aesthetics of the plot  which will be used to specify the axes we are going to use, any groupings, varialbes we are going to colour based on, etc. Unless specifically overwritten in the following layers, all the data information you give in this firts layer will be inheritted in everythiing else you do.

For now we are going to stick with one global specifcation - i.e. we are going to specify all our data in this first layer.

If you run just the first command, R is going to show you the first layer - i.e. a blank screen

Let's try it quickly. Uncomment the comand bellow and see what appears in the 'Plots' section.

```{r}
ggplot()

```

Now let's add some information. Lats give the data and the axes. From our data w ewant to put sex on the x axis and value on the y axis.

What's the difference?

```{r}
ggplot(NULL)

```

*Qiestion: What is the dfference between the above two lines? What does the first layer give you?*


### Layer 2 and onwards

From the second layer onwards, we are specifying how we want our plot to look like. From the type of graph, to the titles and colours. For now we are going to stick with the basics. 



## How do we choose the type of plot?

There are numerous plots we can choose, and we are going to try several today. 

Before plotting we need to know what types of data we want to portray. 

Depending on whether our data is discrete or continuous will change what type of plot is most appropriate.



### Types of data

#### Continuous

Continuous data is that which there are not bounds and the distance between any two consecutivepoints is equivalent to the distance between any other two consecutive points - i.e. the difference between 1 and 2 is the same as the difference between 3 and 4, 5 and 6 etc.  

#### Discrete

Discrete data simply put, groups data. The data in each group or category vary solely on one characteristick and usually those characteristics can not be ranked - i.e. sex, race, nationality etc.


### Number of variables

Next we need to know the relationship between how many variables do we want to show. in a 2D plot we can usually graphically show between 1 and 3 variables with ease. After that more than 2 dimensions are necessary. 

Let's get back to our free movement data.


Q: What types of data are we trying to plot? Continuous or Discrete, or both?

A: 


Q: How many variabes have we chosen so far?

A:


Create a column or a bar plot to show the data use `geom_col()`. How does the data look like?

```{r}

ggplot(NULL) + geom_?

```


Q: Do we need to use another variable to clarify the situation? If yes, which one?

A: 




### Groupings

When we want to group data to provide a better representation, we can usually use colour, shapes etc. Most of the time this will depend on the type of graph we want to use. If we are using line graphs, then colours are usually best. If we are using dot plots then we can use the shape as well as the colours to  differentiate the different groups.

Let's first look at colour.

Depending on the type of graph colouring in can be done either with colour or with fill. For our column plots we use fill.

```{r}
ggplot(traffic_scot, aes(x = Sex, y = Value, fill = `Migration Type`)) + geom_col(position = 'dodge')

```
Let's decompose the code above: 

We have our data and our two variables. They are the same as before. This time we are saying that we want the data to be separated even further by Migration Type. 


Simple task: 

1. Remove `position ='dodge'` from the geom and describe what happens. 


```{r}
#create your plot here

```


We have also added `position = 'dodge'` in the `geom_col()`. This makes sure that any groupings we create and create new columns are then presented separately.

Simple task: 
1. Use `poition = 'fill'` and describe what happens. 


```{r}
#create your plot here

```

The argument `position` is an argument specific for the `geom_col`. It is just one example of an additional argument that helps us to represent the ata better. These types of arguments allow for better control over the plots that we create. 

Geoms in `ggplot2` have a large slecetion of arguments to allow you to adjust your visualisations. This is what makes it one of the  most preffered visualisation packages by the R community. 


#### More groupings

Now lets consider that we want to know whether the individuals are moving to/from Overseas or RUK as well as all the information we have so far. Should we create more groups? Let's try. 

```{r}
ggplot(NULL) + geom_?

```

Here I have used the argument `colour` to give different borders to the migration sources. As you can see this makes it much more difficult to see. Next week we will work on separating the plots.

Q: Would you think there is a better way to represent the data?


A: 

##Now lets have a look at different plots with more complexity

Imagine you want to see the difference in the number of people comming in and out of Scottland from both the RUK and Oveseas throughout the years. 

Q: What Variables would you use?

A:


```{r}
#first lests get the data we will need
# We want our chosen variables, and the overall data for the rest 

traffic_scot2<- NULL

```


Q: What type of data are these variables?

q1: DateCode
q2: Value`
q3: Sex
q4: Migration Source`
q5: Migration Type

```{r}
q1 <- NULL
q2 <- NULL
q3 <- NULL
q4 <- NULL
q5 <- NULL

```

So we need to use a different type of plot - one that can portray continuous data well enough.

Good way to portay data like this are point graphs and line graphs.

We have several varialbes again. It will help if we separate them with different aesthetics.

```{r}
ggplot(NULL) + 
  geom_?

```

This seems difficult to look at. Let's make it easier by adding a line that conects the data. We can do this by addind another *Layer*.


```{r}
ggplot(NULL) + 
  geom_? + 
  geom_?

```


Q: How do we know that the line graph is going to connect the Migration Type and not the Migration Source?


A:



##Practice Task 1

Using the data for this class follow the instructions. 


1. Select two different years. 
2. Select two different age groups. 
3. Select the total for both sex ( i.e. 'all').
4. Select migration In and Out of Scotland.
5. Pick whether you are going to look at 'To-from Resdt of UK' or ' To-from Overseas'.
6. Plot the overall people, regardless of age and year, that have moved into and left Scotland.
7. Plot  the differences between the two years you have selected, regardless of age.
8. Plot the differences between the two ages, regardless of year. 
9. Pick any variables and create a plot that tells a story.


```{r}
# do any wrangling you need to get the data




q6 <- ggplot(NULL) + geom_?
q6

q7 <- ggplot(NULL) + geom_?
q7 

q8 <-ggplot(NULL) + geom_?
q8
```

